---
title: "Exploratory data analysis - School Peformance"
author: "sundar pradeep"
output: html_document
runtime: shiny
---

<br>

#### Synopsis: 

Considering the importance of education in an increasingly knowledge based economy, I performed an exploratory data analysis of school performance in relation to various attributes that might potentially have an influence, with the following objectives. <br>

* Understand what attributes actually influence school performance. <br>
* Analyse if our current affirmative action plans and college admission policies reflect such influence.

<br>

#### Scope, Variables and Datasets: 

Analysis was restricted to `NYC puclic schools` ( comprising 32 school districts)

<br>

#### Factors considered: 

*  `Attendance rate` <br>
*  `School safety`<br>
*  `Class size`<br>
*  `School district size`<br>
*  `Poverty ratio` <br>
*  `Ethnic background` <br>
*  `Gender ratio`<br>
*  `English language learners ratio` <br>
<br>

  `SAT score.` covering Math , Reading and Writing was used as an indicator of school performance.
   
Following datasets were used for the analysis 

1. [School Attendance File](https://data.cityofnewyork.us/Education/School-Attendance-And-Enrollment-By-District-2010-/rfpq-hs49)  
2. [Class size File](https://data.cityofnewyork.us/Education/2009-10-Class-Size-School-level-Detail/d3ge-anaz)    
3. [Demographics File](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=2&cad=rja&uact=8&ved=0CC0QFjAB&url=http%3A%2F%2Fschools.nyc.gov%2FNR%2Frdonlyres%2F2E15D48F-21A5-47CE-984C-208C416EC259%2F0%2FDemographicSnapshot20078to201314Public.xls&ei=6_4GVaLIAca4ggTU84CgCA&usg=AFQjCNFp2A3_mAlfa3w4WuFfHVeoyXfwpQ&sig2=eggltO9pGvESFGMRqhwJ1Q)   
4. [School Safety report](https://data.cityofnewyork.us/Education/School-Safety-Report/qybk-bjjc)   
5. [2010 SAT score file](https://data.cityofnewyork.us/Education/SAT-College-Board-2010-School-Level-Results/zt9s-n5aj)    
6. [2014 SAT score file](http://schools.nyc.gov/Accountability/data/TestResults/default.htm)  


```{r, echo=FALSE, message = FALSE, warning = FALSE }

setwd("~/dev/Projects/1_EDA_NYC_school_data/")

library("dplyr")
library("plyr")
library("tidyr")
library("xlsx")
library("caret")
library("car")
library("leaps")
library("ggthemes")
library("gridExtra")


# 1. load files.
source("./1_load_files.R")

# clean and generate tidy data for analysis.
source("./2_tidy.R")

```


<br>

#### Pre-Processing:

* All source datasets were merged by District-id:School-id to create the master file. 
* The dataset was scaled and centered as the features measured are vastly different- for example , poverty ratio is in percentage , class size in tens and SAT scores in hundreds .
* Data set was checked for Near-Zero variance attributes using `nearZeroVar` function, so they can be dropped from feature set, there were none . 
* Data set was checked for highly correlated variables using `vif` function, so they can be dropped from feature set, there were none . 


```{r,echo=FALSE,  message=FALSE, warning=FALSE}

# model 
# check for Near Zero Variance Predictors
#head(nearZeroVar(district.level.data, saveMetrics = TRUE))
nzv <- nearZeroVar(district.level.data)
#colnames(district.level.data)[nzv] # no near zero variables to drop.


#use vif to check high correlation excl. ethnic background.

subset.data <- district.level.data[,-c(6:9,14)]
lm.subset <- lm(total.percent ~. ,data=subset.data)
vifResult <- vif(lm.subset)

```


<br>

#### Feature Selection: 

`regsubsets` was for used feature selection - following 3 features out of the total 8 feature, were picked up by regsubsets as features that have some influence on SAT scores<br>
`Class size` <br>
`Poverty` and  <br>
`Gender Ratio`

```{r, echo=FALSE, message=FALSE, warning=FALSE}
features.select <- regsubsets(total.percent ~ ., data = district.level.data[,-14])
par(mfrow=c(1,2))
plot(features.select, scale="adjr2")
plot(features.select, scale="bic")
par(mfrow=c(1,1))
```



To cross validate, feature selection was repeated with `steps`  - the same 3 features were picked up by `steps` function as well.


```{r,  echo=FALSE, message=FALSE, warning=FALSE,results='hide' }
# feature selection using steps 

district.model.features <- select(scaled.district.data,-14) 
null=lm(total.percent~1, data=district.model.features)
full=lm(total.percent~., data=district.model.features)
school.step <- step(null, scope=list(upper=full),data=district.model.features, direction="both")
```


```{r,  echo=FALSE, message=FALSE, warning=FALSE }
school.step$coef
```



<br>

#### Influence of Poverty , Class size and gender ratio over school performance:

A linear regression of School Performance with these three variables as the predictors was performed. 


```{r, echo=FALSE, message=FALSE, warning=FALSE}
lm.performance <- lm(total.percent ~ poverty.ratio+size+female.ratio, data=scaled.district.data)
summary(lm.performance)
```
<br>

Only `class-size` and `poverty` features displayed statistically significant influence and hence gender-ratio was dropped from further analysis. 

I decided to take a closer look on the impact of these two key attributes on the overall performance. <br>
When a regression plot on school performance was plotted against class size and poverty, the result was **a surprise**. <br>

<br>

```{r, echo=FALSE,message=FALSE,warning=FALSE}

povertyPlot <- qplot(poverty.ratio,total.percent,data=district.model.features, xlab="Poverty Ratio",ylab="Score",main="Performance Vs Poverty") + 
  geom_smooth(method=lm,formula=y~poly(x,2)) +theme_economist()

sizePlot <- qplot(size,total.percent,data=district.model.features,xlab="Class Size",ylab="Score",main="Performance Vs Class size ") +
  geom_smooth(method=lm,formula=y~poly(x,2)) +theme_economist()

grid.arrange(povertyPlot, sizePlot, ncol=2)
```

<br>

While the influence of poverty on SAT scores was in line with the expectation (increased poverty rates result in decreased scores), the impact of class size was totally unexpected. <br> 
The trend line shows performance declining with smaller class sizes.  

Taking a second look at these plots , the impact of class size over school performance looks like almost a mirror image of poverty plot. I wanted to understand the relation between these two factors. What I found was really interesting.


<br>

```{r,echo=FALSE,warning=FALSE}
qplot(poverty.ratio,size,colour=sizegroup,size=poverty.ratio, data=scaled.district.data,xlab="Poverty Ratio", ylab="class size") + theme_bw()
```


<br>
Clearly, most of the schools in poorer neighborhoods have smaller class sizes than the school districts that are better off. <br>
<br>

This surprising relationship was confirmed by a simple analysis of variance between these two attributes.

<br>

```{r,echo=FALSE,warning=FALSE}
anova(lm(size ~ povertygroup,data=scaled.district.data))

```

<br>
I decided to look for explanation and came across ["Contracts for Excellence Legislation"](http://schools.nyc.gov/AboutUs/funding/c4e/default.htm) .

This legislation funded a set of initiatives over a 5 yr period from 2007, including reduction of class size, focused on poor neighborhoods and schools performing below state standards. 

Our findings confirm that some action seems to have been taken under the legislation since 2007 as reflectef in the 2010 class size data , and that most of the poor neighborhood / low performance shools have comparitively smaller class sizes .<br>

<br>

#### This background presents an interesting question to explore.

Have smaller class sizes really helped to improve the performance of the school districts compared to those with bigger class sizes? <br>
Let us check by comparing the performance of the 2010 class with 2014 class from these schools with smaller class sizes. 
<br>


Let us take a look at performance change of class size groups by comparing 2010 scores against 2014 for each subject (Math , Reading and Writing).

<br>

#### Change in Math performance by class size group:

Linear model and plot comparing the change in math score (2010 vs 2014) is shown below. There seems to be **no visible improvement in the math performance** of schools with smaller classes over this period.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

lm.math.size <- lm(chMath ~ math.sizegroup, data=scaled.school.data)
anova(lm.math.size)

math.plot <- qplot(math.size,chMath,data=scaled.school.data,color=math.sizegroup, xlab="class size - Math",ylab="change in score")+geom_point()+geom_smooth()
math.plot+theme_bw()

```


<br>
<br>

#### Change in Reading performance by class size group:

Linear model and plot comparing the change in reading score (2010 vs 2014) is shown below. There seems to be **no visible improvement in the reading performance** of schools with smaller classes over this period.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

lm.reading.size <- lm(chReading ~ reading.sizegroup , data=scaled.school.data)
anova(lm.reading.size)  

reading.plot <- qplot(english.size,chReading,data=scaled.school.data,color=reading.sizegroup, xlab="class size - Reading",ylab="change in score")+geom_point()+geom_smooth()
reading.plot+theme_bw()
```


<br>

#### Change in Writing performance by class size group:

Linear model and plot comparing the change in writing score (2010 vs 2014) is shown below. There seems to be **no visible improvement in the writing performance** of schools with smaller classes over this period.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

lm.writing.size <- lm(chWriting ~ english.size , data=scaled.school.data)
anova(lm.writing.size)  

writing.plot <- qplot(english.size,chWriting,data=scaled.school.data,color=writing.sizegroup, xlab="class size - Writing",ylab="change in score")+geom_point()+geom_smooth()
writing.plot+theme_bw()
```



Looking at the p-values, there is no confirmation that the smaller class sizes have resulted in a statistically significant improvement in student performance over a three year period.


<br>

#### Conclusion:

* Of all the factors analysed , only poverty seems to have a significant impact in school performance - more than any other factor , including ethnic background. 
Hence , economic background should be an important factor in any affirmative action considerations for college admissions to offset the disadvantage.  <br>

* NYC administration has tried to do its bit by reducing the class sizes to help students from impoverished background , but there is no clear evidence to show that smaller classes have helped improve the performance.<br>
  
<br>

#### References:

Github [Source Code](https://github.com/sundarpradeep/projects/blob/master/EDA_School_Performance.Rmd)

<br>
<br>

